---
title: "the basic rmd 2"
author: "Adrianna Pops"
date: "2024-09-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Description

* Variable 1: rank - a 3 level job ranking for an individual that includes Associate Professor, Assistant Professor, or Professor, character
* Variable 2: discipline - whether the individual is in discipline A or discipline B, character
* Variable 3: yrs.since.phd - years since the individual has completed their phd in years
* Variable 4: yrs.service - how long the individual has been employed at this university in years
* Variable 5: sex - whether the individual is male or female, character
* Variable 6: salary - the salary the individual earns, numeric

These are the variables for the data set salary. They included job rank, discipline, years since phd, years of service, sex and salary earned.

# Data Wrangling

* Step 1: I installed the required packages (knitr, reader, tidyverse) using install.packages(). I already have knitr and readr installed locally so I only installed tidyverse for this lab.
* Step 2: I loaded tidyverse using library().
* Step 3: I set my working directory using setwd()
* Step 4: I read the csv file Salaries using the read_csv() function
* Step 5: I viewed Salaries to check variables and data types using view()
* Step 6: I selected columns rank, discipline, sex, and salary using select()
* Step 7: I filtered out missing salaries using filter()
* Step 8: I grouped the data by rank, discipline, and, sex using group_by()
* Step 9: I tallied for each group using count()
* Step 10: I combined steps 6 - 9 using a pipe operator ( %>% ) and assigned the output to a new variable.
* Step 11: I used spread() to create three new columns for "Male", "Female", and the count column "n"
* Step 12: I used gather() to reshape the table back to long and combined the "Male" and "Female" columns back to sex.

These are the steps I used from previous modules in order to take raw data and analyze/visualize it.

```{r, results=FALSE, message=FALSE}
#install tidyverse package
#install.packages("tidyverse")

# 1. Load the library
library(tidyverse)

# 2. Set the working directory
setwd("/Users/ampops/Desktop/dsci605/module3")

# 3. Read csv file into R using the function read_csv().
Salaries <- read_csv("/Users/ampops/Desktop/dsci605/module2/Salaries.csv")
View(Salaries)

# 4. Use select(), filter(), group_by(), count() in this part and combine them with pipe operator
# + Assign the output to a new object named "Salaries_new." "Salaries_new" should have four columns now. 
# + Order the steps by
#   - Select some columns
#   - Filter the data where "salary" is not missing value.
#   - Group the data based on the columns of "rank", "discipline", "sex"
#   - Get the tally count of the observation

# Used a pipe to combine the select(), filter(), group_by(), and count() steps and assigned the output to the variable Salaries_new
Salaries_new <- Salaries %>% 
  select(rank, discipline, sex, salary) %>% 
  filter(salary != 0) %>% 
  group_by(rank, discipline, sex) %>% 
  count(rank, discipline, sex)

view(Salaries_new)

# 5. using spread() to transfer the "sex" column to many more columns based on the values in sex.
# Use spread(): spread(data, key, value, fill = NA, convert = FALSE, drop = TRUE, sep = NULL)
# + Please use the data frame "Salaries_new" for the spread() operation
# + Please split the column "sex" in the data frame "Salaries_new" to many more columns based on the values in sex
# + Have the values of n(tally count) for the new columns.
# + Fill the missing value with "FALSE" by setting "fill=FALSE"
# + Assign the output to an object "wide"
# + Use pipe operator in this case

# Used spread() to create three new columns for "Male", "Female" and "n" which is the tally of the columns.
wide <- Salaries_new %>%
  spread(sex, n, fill = FALSE, convert = FALSE, drop = TRUE, sep = NULL)

view(wide)

# 6. Reshape the table from wide to long: using gather() to combine multiple "sex" related columns to one column sex
# Use gather(): gather(data, key, value, ..., na.rm = FALSE, convert = FALSE, factor_key = FALSE)
# + Please reshape the data frame "wide" obtained from spread() into the long table 
# + In the spread() step, you will get two columns about sex "Female" and "Male"
# + Please make the two sex columns into one column and name the column as "sex"
# + Please have the values in the two sex columns for a new column "n" (tally count)
# + Remove the NA by setting "na.rm=TRUE"
# + Use pipe operator in this case

# Used gather() to recombine the sex column values and reshape the table back to long. 
long <- wide %>% 
  gather(sex, n, Female, Male, na.rm = TRUE, convert = FALSE, factor_key = FALSE)

view(long)
```

